Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-30 draft.tex
Included Bernhards feedback

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate difference between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques, which reach beyond standard statistical methods.

@@

The standard workflow lacks methods, which identifies correlation the epidemiologists did not consider.

@@

%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
In this section we describe the epidemiological workflow and associated requirements.
% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
%
%Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors, to medical computer scientists with focus on biometrics, and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it doesn't consider any kind of image data.
%
The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
%In the following we discuss the problems and challenges regarding the workflow.

@@

In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20-year-steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tisse).

@@

\section{Prior and Related Work}
%Einfuehren von Helwigs Terminologie?
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measure highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and derive variable associations with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Derived from this, we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference \cite{Hermann2014}. 
%
The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values.

@@

Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues.

@@

Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop} seen in Figure~\ref{fig:WorkflowComparison}. Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types. The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new features can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
The visualization type is, inspired by \texttt{GPLOMS} \cite{Francois2013}, dynamically chosen based on the feature types and number.
%
Categorical data are mapped either using bar charts (single features) or mosaic plots (multiple features).
%
Continuous data are discretized and therefore treated as categorical data.
%
It is, however, also possible to plot it using scatterplots (single features) or parallel coordinates (multiple features).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%Inspired by the previously discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
%
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D-plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
%
Mapping the distance from a group mean shape to the global mean to color allows to assess local shape changes.
%%TODO Adaptive visualization!
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
%
Each plot can be brushed using widgets.
%
Brush selections are propagated to all visualizations allowing for fast feature querying.
% TODO Timeline
%
%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
%
%A use case for this is when a continuous feature has to be divided into even groups.

@@

\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
% TODO: Has to be triggered since visualization is very large
As discussed previously, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA} powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape-space.
%
Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
The \emph{Cram\'{e}rs' V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}rs' V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total of observations and $k$ either the row or column count, depending on which one is lower.
%
$V$ assumes values between $0$, meaning two variables are completely independent, and $1$, indicating they are the same.
%
\emph{Cram\'{e}rs' V} is always positive and does therefore not allow statements about dependency direction.

>>>>

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are used to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques, which reach beyond standard statistical methods.

@@

The standard workflow lacks methods, which identify correlations the epidemiologists did not consider.

@@

%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
In this section we describe the epidemiological workflow and associated requirements.
% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
%
%Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors, to medical computer scientists with focus on biometrics, and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it does not consider any kind of image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.
%The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
%In the following we discuss the problems and challenges regarding the workflow.

@@

In longitudinal cohort studies, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20-year-steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.

@@

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks labeling the voxels of an anatomical structure would be ideal since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and number of subjects.
%
Information derived by landmarks, such as top and bottom point are by far not as expressive and versatile as segmentation masks describing the whole vertebra.
%Information derived by landmarks such as top and bottom point of a vertebra are also not nearly as expressive and versatile as segmentation masks describing the whole vertebra.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
%This gains even more momentum when analyzing multiple time steps.
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey-value distribution in an area (used for determining the type of tisse).

@@

\section{Prior and Related Work}
%Einfuehren von Helwigs Terminologie?
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on \emph{hypothesis generation} using parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

%\emph{Over-fitting} of hypotheses to the data may be imposed  is a danger pointed out by the authors when hypotheses to the data based on observations of changes in these plots the measure highlights only subsets of statistical changes.
Hypotheses based on observations of changes in these plots may impose \emph{overfitting} to the data because the measures highlights only subsets of statistical changes.
%
Our approach sticks more to the information extracted from the segmented image data and variable associations with non-image epidemiological factors.
% Disadvantage - Nur 2 abgeleitete Variablen sichtbar, schlecht mit anderen Variablen in Verbindung zu bekommen (müsste jedoch auch gehen) - viele Dinge gehen durch die Abstraktion aber auch flöten!

@@

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable a Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems with their origin in business intelligence applications do not support epidemiological workflows, in particular workflows that also consider medical image data.
%These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
We incorporate the idea of combining 3D shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference on rodent mandibles \cite{Hermann2014}.
% TODO: Blöde formulierung
Specifying a change in shape of a certain area by the user shows how other parts of the model would react using covariance tensors.
%Covariance tensors indicate how a model is deformed when the user given a user specified deformation.
%The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

@@

A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables, the meaning and unit of measurement of their values.

@@

Missing data are denoted using error codes indicating its cause ranging from ethical to medical and personal issues (recall Subsection~\ref{EpidemiologicalData}).

@@

%Gerade weil der Satz davor so kompliziert war, ist unklar, was "this" bedeutet. Ich denke zunächst muss gesagt werden, was typische Ziele sind (anhand von Beispielen, wie Abstände zwischen anatomischen Landmarken, Ausdehnungen (also längste Durchmesser). Wichtig wären aber konkrete relevante Beispiele. Dann wird gesagt, was man dafür machen muss und wie man das macht.
Information about anatomical structure, such as diameter or volumes, is extracted from the image data.
%Processing the image data associated to each subject consists for the most part of information extraction about an anatomical structure.

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed \emph{IVA Loop} as seen in Figure~\ref{fig:WorkflowComparison}. Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types. The data are visualized in the range and domain space, which can then be brushed, yielding new groups, which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The Sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{age} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis-generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new features can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features. %and lead therefore to a new feature selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\paragraph{Adaptive Feature Visualization.} \label{sec:AdaptiveFeatureVisualization}
The visualization type is, inspired by \texttt{GPLOMS} \cite{Francois2013}, dynamically chosen based on the feature types and number.
%
Categorical data are mapped either using bar charts (single features) or mosaic plots (multiple features).
%
Continuous data are discretized and therefore treated as categorical data.
%
It is, however, also possible to plot it using scatterplots (single features) or parallel coordinates (multiple features).
%
Figure \ref{fig:visualization} describes this dynamic adjustment.
%Inspired by the previously discussed \texttt{GPloms} \cite{Francois2013} the visualization type is chosen dynamically based on the variable types and number of visualizations which need to be displayed.
%
% TODO Implement MOSAIC PLOTS, Scatterplots, Parallel Coordinates
Following Tuftes concept of \emph{small multiples} \cite{Tufte1983}, the medical image data are directly incorporated into the plot by including color-coded mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%If possible the medical image data is directly included into the plot as well by including mean shapes for each manifestation (Figure~\ref{fig:visualization} (b)).
%
The 3D plots can be navigated using standard mouse inputs, the camera is synchronized between all views to enable direct comparison.
%
Mapping the distance from a group mean shape to the global mean to color allows to assess local shape changes.
%%TODO Adaptive visualization!
If a feature is dropped on an existing plot, the visualization changes dynamically to properly make them comparable. 
%
Each plot can be brushed using widgets.
%
Brush selections are propagated to all visualizations allowing for fast feature querying.
% TODO Timeline
%
%It is possible to duplicate brushes in order to create new groups which are evenly spread out.
%
%A use case for this is when a continuous feature has to be divided into even groups.

@@

\paragraph{Automated Feature Suggestion using a Contingency Matrix.}
% TODO: Has to be triggered since visualization is very large
As discussed previously, highlighting potentially interesting values in the data set is one major benefit of the \emph{IVA} powered approach and belongs to the \emph{multivariate analysis} pattern,  analyzing features outside the shape-space.
%
Turkay and colleagues used the approach to calculate various key figures based on the distribution functions of each feature derived from the image data \cite{Turkay2013}.
%
Since the majority of our data are categorical features, we have to employ different solutions.
%
The \emph{Cram\'{e}rs' V} contingency coefficient can be used to calculate coherences between categorical variables \cite{CramerV}.
%
It is based on \emph{Pearson's $X^2$} distribution test \cite{ChiSquare}, which uses contingency tables holding the counts of subjects for all possible manifestations of two variables.
%
\emph{Cram\'{e}rs' V} is defined as:
\begin{equation}
V = \sqrt{\frac{X^2}{N(k-1)}},
\end{equation}
where $X^2$ equals \emph{Pearson's chi squared}, $N$ is the total number of observations and $k$ either the row or column count, depending on which one is lower.
%
$V$ assumes values between $0$, meaning two variables are completely independent, and $1$, indicating they are the same.
%
\emph{Cram\'{e}rs' V} is always positive and does therefore not allow statements about dependency direction.
